\section{Additional Analysis: Synchronization Dynamics}
\label{sec:additional}

We analyze the \emph{synchronization dynamics} of the transformer: how quickly and through which mechanism does the model identify the active component? This analysis is motivated by the view that in-context learning is fundamentally source discrimination in non-ergodic data.

\subsection{Motivation}

In the non-ergodic mixture, the model faces a hierarchical inference problem at each sequence:
\begin{enumerate}
    \item \textbf{Meta-synchronization}: Determine which Mess3 component generated this sequence.
    \item \textbf{Within-component belief tracking}: Given the component, track the belief state over hidden states.
\end{enumerate}
These two tasks have different computational requirements. Meta-synchronization requires comparing accumulated token statistics against learned priors for each component---a multiplicative (likelihood-ratio) computation in log-probability space. Within-component belief tracking requires the constrained Bayesian update described in~\cite{piotrowski2025constrained}.

We probe each task independently to understand when and where each representation emerges.

\subsection{Meta-Belief Accuracy by Position}

We train separate linear probes at each context position $t$ to predict (a) the meta-belief $P(k \mid z_{1:t})$ and (b) the within-component belief $P(s \mid k, z_{1:t})$ from the final-layer residual stream.

\textit{[Results to be filled: R$^2$ curves showing meta-belief accuracy increasing with position, expected to be faster for well-separated components.]}

\subsection{Layer-Wise Probing}

At each layer, we independently probe for component identity and within-component belief state.

\textit{[Results to be filled: R$^2$ at each layer for meta-belief vs within-component belief. Expected: meta-belief emerges earlier than belief refinement.]}

\subsection{Attention Pattern Analysis}

We examine the average attention patterns across layers to determine whether specific heads or positions are specialized for meta-synchronization.

\textit{[Results to be filled: attention heatmaps. Expected: early-layer heads attend broadly (gathering evidence for source discrimination), later-layer heads attend more locally (for belief updating).]}

\subsection{Connection to In-Context Learning}

The synchronization dynamics we observe provide a concrete mechanistic account of in-context learning: the model first identifies the active source (analogous to topic/style identification in natural language), then conditions its predictions on this inferred source identity. This two-phase structure---source identification followed by conditioned prediction---may be a universal computational motif in sequence models trained on non-ergodic data.

The per-position loss curve, when compared against the Bayes-optimal baseline, quantifies the ``synchronization gap'': the additional prediction error incurred by imperfect source identification and belief tracking.
