\section{Discussion}

\paragraph{Non-ergodicity as the heart of in-context learning.} The non-ergodic Mess3 mixture provides a minimal, analytically tractable model of in-context learning. The key insight is that non-ergodicity \emph{forces} the model to perform hierarchical inference: it cannot achieve low loss without first identifying the source. This is precisely the situation language models face---they must identify the ``regime'' (topic, style, task, speaker) before they can predict well. Our setup strips this to its mathematical essence.

\paragraph{Block-diagonal vs tensor-product structure.} The joint transition operator of our mixture is block-diagonal, not a tensor product. This means the standard FWH does not directly apply. Yet we observe that the model still learns a partially factored representation, suggesting that the factoring inductive bias extends beyond the conditionally-independent case. This is consistent with the FWH acting as a ``representational attractor'' that models approach even when exact factoring is not achievable.

\paragraph{Residual stream as belief space.} The residual stream naturally serves as a space for representing beliefs. Since it has no privileged basis, the geometric arrangement of activations encodes similarity of predictions: nearby activations make similar forecasts about the future. The factored structure we observe means the model has discovered that some dimensions of prediction (source identity) are independent of others (within-source state), and has allocated orthogonal subspaces accordingly.

\paragraph{Limitations and future work.}
\begin{itemize}
    \item Our model is small (3--4 layers, 64--128 dimensions). Scaling to larger models could reveal whether the geometric structures become sharper or qualitatively different.
    \item Our components are static within sequences. Slowly drifting or switching parameters would test how the model handles non-stationary non-ergodicity.
    \item We use the same token alphabet across components. A natural extension is to consider processes with shared and component-specific tokens, creating a richer factorization structure.
    \item Connecting the attention-based meta-synchronization mechanism to induction heads \cite{olsson2022context} could reveal whether the same circuits serve both in-context learning tasks.
\end{itemize}
